### 2021-07-06 rad-icfp-2021/intro.tex
.

\subsection{Motivation}

In 2008, Pearlmutter and Siskind~\cite{pearlmutter2008reverse} introduced Automatic Differentiation (AD) in a functional setting, where they defined reverse-mode. 
Reverse-mode computes in a single fast pass the whole gradient of a function $\RR^n\to\RR$.
By comparison, the easier forward-mode transformation would require $n$ passes. 
These functions $\RR^n\to\RR$ appear a lot in practice: for instance, as loss functions in machine learning.
Reverse-mode is thus important, as it computes the whole gradient of such functions very quickly. 
However, in order to reach the efficiency of the usual imperative version of reverse-mode, the transformation in~\cite{pearlmutter2008reverse} resorts to references.

The lack of purity in reverse-mode is known to make it significantly harder to optimize and parallelize. 
None of the current implementations of reverse-mode in a functional setting~\cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.~\cite{xla}.

As a result, to optimize for efficiency, a hand-crafted reverse-derivative must be given, for every important non trivial operation.
In other words, abstracting away from imperative code is still a hurdle that functional implementations need to overcome.

An efficient purely functional reverse-mode transformation would be more easily explainable, optimizable, and reusable.

TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g. in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress. 
For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.

\subsection{Problem}

Following \cite{pearlmutter2008reverse}, it is relatively easy to define a purely functional reverse-mode on a first-order language using continuations. 
Yet, as remarked by the authors, this is highly inefficient. 
The problem we tackle in this paper is how to go from a simple and general setting, where we can relatively easily prove correctness of reverse-mode, 
to a representation that leads to more efficient implementations.

@@

>>>>

% \subsection{Motivation}
Deep learning is moving towards increasingly sophisticated optimization objectives that employ tensors and operations on tensors.
Reverse-mode Automatic Differentiation (AD) is a technique to automatically compute the gradient of objective functions of form $\RR^n\to\RR$.
Such functions appear a lot in practice: for instance, as loss functions in machine learning.

In order to reach the efficiency of the usual imperative version of reverse-mode, the transformations even in functional languages~\cite{pearlmutter2008reverse} resort to references.
The lack of purity in reverse-mode is known to make it significantly harder to optimize and parallelize. 
None of the current implementations of reverse-mode in a functional setting~\cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g.~\cite{xla}.
As a result, to optimize for efficiency, a hand-crafted reverse-derivative must be given, for every important non-trivial operation.
In other words, abstracting away from imperative code is still a hurdle that functional implementations need to overcome.

% An efficient purely functional reverse-mode transformation would be more easily explainable, optimizable, and reusable.

% TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
% all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g. in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress. 
% For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.

% \subsection{Problem}

% Following \cite{pearlmutter2008reverse}, it is relatively easy to define a purely functional reverse-mode on a first-order language using continuations. 
% Yet, as remarked by the authors, this is highly inefficient. 
% The problem we tackle in this paper is how to go from a simple and general setting, where we can relatively easily prove correctness of reverse-mode, 
% to a representation that leads to more efficient implementations.

In this paper, we define a purely functional language with efficient and provably correct reverse-mode AD.
To do so, we go start from a simple and general representation, where we can relatively easily prove correctness of reverse-mode.
Then, we define a representation that leads to more efficient implementations.

@@

% \begin{table}
% \begin{tabular}{|l|c|c|c|c|c|c|c|c|c|}
% \hline
%  & \rot{Reverse Mode}  & \rot{Complexity Proofs} & \rot{Pure} & \rot{Correctness Proofs} & \rot{Tensor Support}  & \rot{Recursion} & \rot{Conditional} \\
% \hline
% This Paper &
% \supfull & \supfull & \supfull  & \supfull  & \supfull  & \suphalf & \supfull \\ \hline
% Lantern~\cite{lantern_icfp} & 
% \supfull & \suphalf & \supnone  & \supnone & \supfull & ? & \supfull \\ \hline
% \dfsmooth{}~\cite{shaikhha2019efficient} & 
% \supnone & \supnone & \supfull  & \supnone & \supfull  & \suphalf  & \supnone \\ \hline
% \cite{huot2020correctness} &
% \supfull & \supnone & \supfull   & \supfull & \supnone & \suphalf & \supfull \\ \hline
% \cite{brunel2019backpropagation} &
% \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & ? \\ \hline
% \cite{abadi2019simple} &
% \supfull & ? & \supfull & \supfull & \supnone & \supfull & \supfull \\ \hline
% \cite{barthe2020versatility} &
% \supnone & \supnone & \supfull & \supfull & \supnone & \supnone & \supfull \\ \hline
% \cite{pearlmutter2008reverse} &
% \supfull & \suphalf & \supnone & \supnone & ? & ? & ? \\ \hline
% \cite{Elliott:2018:SEA:3243631.3236765} &
% \supfull & \suphalf & \supfull & \supnone & ? & \supnone & ? \\ \hline
% \end{tabular}
% \caption{Comparison of different functional differentiable programming frameworks.}
% \end{table}

\begin{table}
 \label{fig:comparison-table}
 \begin{tabular}{|l|c|c|c|c|c|c|c|c|}
 \hline
  & \rot{Reverse Mode}  & \rot{Complexity Proofs} & \rot{Pure} & \rot{Correctness Proofs} & \rot{Tensor Support}  & \rot{Recursion} & \rot{Conditional} \\
  % & Reverse  & Complexity & Pure & Correct & HO Fun. & Tensors & Rec. & Cond. \\
 \hline
 \system{} (This Paper) &
 \supfull & \supfull & \supfull & \supfull  & \supfull  & \supfull & \supfull \\ 
 \hline
 Lantern~\cite{lantern_icfp} & 
 \supfull & \suphalf & \supnone & \supnone & \supfull & \supfull & \supfull \\ 
 \hline
 \dfsmooth{}~\cite{shaikhha2019efficient} 
 & 
 \supnone & \supnone & \supfull & \supnone & \supfull  & \supnone  & \supnone \\ 
 \hline
 \cite{huot2020correctness} &
 \supfull & \supnone & \supfull & \supfull & \supnone & \suphalf & \supfull \\ 
 \hline
 \cite{brunel2019backpropagation} &
 \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
 \hline
 \cite{abadi2019simple} &
 \supfull & \supfull? & \supfull & \supfull & \supnone? & \supfull & \supfull \\ 
 \hline
 \cite{barthe2020versatility} &
 \supnone & \supnone & \supfull & \supfull  & \supnone & \supnone & \supfull \\ 
 \hline
 \cite{pearlmutter2008reverse} &
 \supfull & \supnone  & \supnone & \supnone & \supnone? & \supfull & \supfull \\ 
 \hline
 \cite{Elliott:2018:SEA:3243631.3236765} &
 \supfull & \suphalf & \supfull & \supnone & \supnone? & \supnone & \supnone? \\ 
 \hline
 \cite{sherman2021} & 
 \suphalf or \supnone? & \supnone & \supfull & \supfull? & \supnone? & \supnone & \supnone \\ 
 \hline
 \cite{vytiniotis2019differentiable} &
 \supfull & \suphalf & \supfull & \supnone & \supnone? & \supnone & \supnone? \\ 
 \hline
 \cite{mak2020differential} & 
 \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
 \hline
 \cite{vakar2020reverse} & 
 \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
 \hline
 \cite{Manzyuk2012} & 
 \supnone & \supnone & \supfull & \supfull? & \supnone & \supnone & \supnone \\ 
 \hline 
 \cite{gallagher-sdg} & 
 \supnone & \supnone & \supfull & \supfull ? & \supnone & \supfull & \supfull  \\ 
 \hline
 \end{tabular}
 \caption{Comparison of different differentiable programming frameworks.
 $\supfull$ means the property is verified, and $\supnone$ means its absent from the work.}
 \end{table}
